\chapter{Examples}

This Section presents four choreographic programs, to demonstrate how choreographies can be written in Lean using the chorlean library. The first example \ref{ss:silent} shows a very basic communication scheme. In the \ref{ss:book} example a already more complex protocol is presented to highlight higher Level Choreographies aswell as Location Polymorphism. \ref{ss:sso} Shows an application of a simple cryptographic protocol for a SSO service and the last example \ref{ss:merge} showcases a compute heavy application that balances compute load of a sorting algorithm across a set of  nodes.

\subsection{silent post}
\label{ss:silent}
This first Distributed protocol is kept simple to demonstrate some of the features(TODO einen richtigen Satz formulieren).
Three Participants, Alice, bob and eve take part in this protocol. The Scheme is that alice prompts an text input from the user, Then passes this input arround in a circle through bob and eve. Both bob and eve perfom local modifications of the message in the spirit of the silent post game.
%\input{chapters/examples/silentpost}
As the Type of a Choreo depends on the endpoint it is projected to, the silent_post function carries an endpoint as a argument. Also notice the \emph{@} notation, that is an abbreviation for an \emph{GVal} Type with implicit endpoint. This notation works inside a Choreo as the endpoint of GVal's resulting from a communication or local operation is allways the endpoint the choreography is parameterized with.
As discussed in --TODO-- the locally functions lets uns define a Freer-Monad with all locally available effects. For Silent post, those local Effect Signatures are defined as
\begin{lstlisting}[language=lean]
instance sig: LocSig Location where
  sig x := match x with
    | alice =>  CmdInputEff ⨳ LogEff
    | bob =>  LogEff
    | eve => LogEff
\end{lstlisting}
such that every location is allowed to perform a logging effect \emph{LogEff} and additionally alice has access to CmdInputEff, which allow to prompt a String from a User in the command line.


\subsection{book seller}
\label{ss:book}
\input{chapters/examples/bookseller}

\subsection{sso authentication}
\label{ss:sso}

\input{chapters/examples/sso}
\subsection{mergesort}
\label{ss:merge}

%\input{chapters/examples/mergesort}
